---
title: "Homework2"
author: "Yahan Xue"
date: "2026-10-04"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

# Problem 1

```{r}
subway_df = 
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = entry == "YES"
  )
```

```{r}
subway_df |>
  distinct(station_name, line) |>
  nrow()
```

```{r}
subway_df |>
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```

```{r}
subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion_entry = mean(entry)
  )
```

```{r}
route_df = 
  subway_df |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    values_drop_na = TRUE
  )

route_df |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()

route_df |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(station_name, line) |>
  nrow()
```

The cleaned subway dataset contains 1868 observations and 19 variables, including station name, line, station latitude and longitude, routes served, entry status, vending availability, entrance type, and ADA compliance. The entry variable was converted from character values to a logical variable. The original route variables are not tidy because route information is spread across multiple columns. There are 465 distinct stations, of which 84 are ADA compliant. Among entrances and exits without vending, approximately 37.7% allow entrance. After reorganizing the route variables, 60 distinct stations serve the A train, and 17 of these stations are ADA compliant.

# Problem 2

```{r}
library(readxl)

mr_trash = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.numeric(year),
    month = str_trim(month),
    plastic_bottles = readr::parse_number(as.character(plastic_bottles)),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )

professor_trash = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.numeric(year),
    month = str_trim(month),
  plastic_bottles = suppressWarnings(
  readr::parse_number(as.character(plastic_bottles))
),
trash_wheel = "Professor Trash Wheel"
  )

gwynnda_trash = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394"
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.numeric(year),
    month = str_trim(month),
    plastic_bottles = readr::parse_number(as.character(plastic_bottles)),
    trash_wheel = "Gwynnda"
  )

trash_wheel_df = 
  bind_rows(
    mr_trash,
    professor_trash,
    gwynnda_trash
  )

nrow(trash_wheel_df)

trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  )

trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE)
  )
```

The combined Trash Wheel dataset contains 1269 observations. Key variables include dumpster number, month, year, weight of trash collected in tons, volume, plastic bottles, cigarette butts, and the Trash Wheel identifier. The data were imported from the three relevant Excel sheets, cleaned by removing non-data rows, and combined into one dataset. The total weight of trash collected by Professor Trash Wheel was 294.96 tons. In June 2022, Gwynnda collected a total of 18,120 cigarette butts.

# Problem 3
```{r}
mci_baseline = 
  read_csv(
    "data/MCI_baseline.csv",
    skip = 1,
    na = "."
  ) |>
  janitor::clean_names() |>
  mutate(
    sex = if_else(sex == 1, "Male", "Female"),
    apoe4 = if_else(apoe4 == 1, "Carrier", "Non-carrier")
  ) |>
  filter(
    is.na(age_at_onset) | age_at_onset > current_age
  )

mci_baseline
```

```{r}
nrow(mci_baseline)

mci_baseline |>
  filter(!is.na(age_at_onset)) |>
  nrow()

mean(mci_baseline$current_age)

mci_baseline |>
  filter(sex == "Female") |>
  summarize(
    proportion_apoe4 = mean(apoe4 == "Carrier")
  )
```

```{r}
mci_amyloid = 
  read_csv(
    "data/mci_amyloid.csv",
    skip = 1,
    na = c("NA", "Na")
  ) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    cols = baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio"
  ) |>
  mutate(
    time = case_when(
      visit == "baseline" ~ 0,
      visit == "time_2" ~ 2,
      visit == "time_4" ~ 4,
      visit == "time_6" ~ 6,
      visit == "time_8" ~ 8
    )
  ) |>
  select(id, time, amyloid_ratio)

mci_amyloid
```

```{r}
baseline_only = 
  mci_baseline |>
  anti_join(
    mci_amyloid |> distinct(id),
    by = "id"
  )

amyloid_only = 
  mci_amyloid |>
  distinct(id) |>
  anti_join(
    mci_baseline,
    by = "id"
  )

baseline_only$id

amyloid_only$id
```
```{r}
mci_combined = 
  inner_join(
    mci_baseline,
    mci_amyloid,
    by = "id"
  )

mci_combined

n_distinct(mci_combined$id)
dim(mci_combined)
```

```{r}
write_csv(
  mci_combined,
  "data/mci_combined.csv"
)
```
Following cleaning of the baseline demographic dataset, 479 participants satisfied the study’s inclusion criteria. Of these individuals, 93 progressed to MCI over the follow‑up period. The mean baseline age was roughly 65.03 years, and 30% of female participants were APOE4 carriers.

The amyloid biomarker dataset was imported and converted from wide format to long format, such that each row corresponds to one participant for a single study visit. Cross‑comparison of participant IDs between the two datasets revealed 8 participants present exclusively in the baseline dataset and 16 participants found only within the amyloid biomarker dataset. An inner join was subsequently performed, yielding a final sample of 471 participants common to both datasets. This merged dataset integrates baseline demographic data with longitudinal amyloid measurements and was saved and exported in CSV file format.